---
title: "Vignette for NPA Data Acquisition and Organization"
author: "Daivic Akala"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview
This vignette demonstrates how to use the NPADataAcqOrg() function from the NetworkPropertyAnalysis package. Specifically, the NPADataAcqOrg() function processes raw Xenium output transcript data in the form of a CSV file, selects and organizes relevant columns (cell_id, feature_name, x_location, y_location), saves both the full grouped data set and a sampled subset to determine additional parameters, and builds a hierarchal directory by Cell and Gene, saving individual .csv files for each identified gene within each sampled cell

# Parameters
```{r}
#' @param TranscriptFile Describes the file path to the CSV file that contains the raw transcript data from Xenium
#' @param OutputDir Directory created to save intermediate CSV files
#' @param GroupedPath Describes the file path to the CSV file that contains the grouped information from the GroupedData variable
#' @param GroupedSamplePath Describes the file path to the CSV file that contains the grouped information from the SampleData variable
#' @param SampleSize Number of cells to sample for Distance Threshold Determination (Default 1000)
#' @param Seed Random seed value for reproducibility (Default 123)
```

This section of the code reads the transcript-level spatial data from a CSV file into a data.table using fread(), which is optimized for fast I/O. It then selects only the relevant columns (cell_id, transcript_id, feature_name, x_location, and y_location) and ensures the result is a data.table. The data is sorted by cell_id and feature_name to prepare for downstream organization. If a valid GroupedPath is provided, the filtered and sorted data is saved to that file using fwrite(). Finally, the table is keyed on cell_id and feature_name, which enables efficient subsetting and joins in later steps of the pipeline.

```{r}
TranscriptData <- data.table::fread(TranscriptFile)
  SelectedData <- data.table::as.data.table(TranscriptData)[, c("cell_id", "transcript_id", "feature_name", "x_location", "y_location"), with = FALSE]
  GroupedData <- SelectedData[order(cell_id, feature_name)]
  if (!is.null(GroupedPath)) {
    data.table::fwrite(GroupedData, GroupedPath, row.names = FALSE)
  }
  data.table::setkey(data.table::setDT(GroupedData), cell_id, feature_name)
```

This section of the code performs random sampling and prepares the output directory for organized storage. First, it sets the random seed using `set.seed(Seed)` to ensure reproducibility of the sampled cells. It then randomly selects a subset of unique `cell_id`s from the grouped data, equal in number to `SampleSize`. The transcripts corresponding to these sampled cells are extracted into `SampledData`. If a file path is provided for `GroupedSamplePath`, the sampled subset is saved as a CSV using `fwrite()`. Finally, the output directory specified by `OutputDir` is created (if it doesn’t already exist), allowing subsequent data files to be stored in an organized folder structure.

```{r}
  set.seed(Seed)
  SampledCells <- sample(unique(GroupedData$cell_id), size = SampleSize)
  SampledData <- GroupedData[cell_id %in% SampledCells]

  if (!is.null(GroupedSamplePath)) {
    data.table::fwrite(SampledData, GroupedSamplePath, row.names = FALSE)
  }

  dir.create(OutputDir, recursive = TRUE, showWarnings = FALSE)
```

This section organizes and saves the sampled transcriptomic data into a structured folder hierarchy. It begins by identifying all unique cell IDs (`UniqueCells`) from the `SampledData` and calculating the total number of sampled cells (`nCells`). For each cell, it extracts the subset of data (`CellDataSubset`) corresponding to that cell and creates a directory named `Cell_<cell_id>` within the specified `OutputDir`. Inside each cell's directory, it iterates through all unique genes (`UniqueGenes`) expressed in that cell. For each gene, it extracts the corresponding transcripts (`GeneData`), creates a subdirectory `Gene_<gene>`, and writes the gene-specific data to a uniquely named CSV file. Throughout the loop, progress is displayed in the console using `cat()` and `flush.console()`, showing the percentage of cells processed. This structured output is essential for downstream analyses that depend on per-cell and per-gene data organization.

```{r}
UniqueCells <- unique(SampledData$cell_id)
  nCells <- length(UniqueCells)

  for (i in seq_along(UniqueCells)) {
    Cell <- UniqueCells[i]
    CellDataSubset <- SampledData[cell_id == Cell]
    CellDir <- file.path(OutputDir, paste0("Cell_", Cell))
    dir.create(CellDir, showWarnings = FALSE, recursive = TRUE)

    UniqueGenes <- unique(CellDataSubset$feature_name)

    for (gene in UniqueGenes) {
      GeneData <- CellDataSubset[feature_name == gene]
      if (nrow(GeneData) == 0) next
      GeneDir <- file.path(CellDir, paste0("Gene_", gene))
      dir.create(GeneDir, showWarnings = FALSE, recursive = TRUE)
      OutputFile <- file.path(GeneDir, paste0("Gene_", gene, "_Cell_", Cell, "_Data.csv"))
      data.table::fwrite(GeneData, OutputFile)
    }

    #Progress Reporting
    ProgressPercent <- (i / nCells) * 100
    cat(sprintf("Directory Creation Progress: %.1f%% \r", ProgressPercent))
    flush.console()
  }
```

# Sample Function Call
```{r}
NPADataAcqOrg(
 TranscriptFile = "C:/Users/daivi/Documents/Daivic/Summer 2025/IBRI/ND16090 T2D 0017308/transcripts.csv",
  OutputDir = "C:/Users/daivi/Documents/Daivic/Summer 2025/IBRI/ND16090 T2D 0017308/Package Test/CellGeneDirectories",
  GroupedPath = "C:/Users/daivi/Documents/Daivic/Summer 2025/IBRI/ND16090 T2D 0017308/Package Test/Grouped Data.csv",
  GroupedSamplePath = "C:/Users/daivi/Documents/Daivic/Summer 2025/IBRI/ND16090 T2D 0017308/Package Test/Grouped Sample Data.csv",
  SampleSize = 1000,
  Seed = 123
)
```

